---
title: "Assignment RR 1 Week 2"

output: md_document



---

This is the assignment which processes and analyzes personal monitoring activity data.

In order for this to run smoothly, you need to install packages DoBy and broman!

```{r,echo=TRUE, warning=FALSE, message=FALSE}
library("doBy")
library(broman)
library(lattice)
```


1) Retrieving the data and checking if variables are in proper format.

```{r,echo=TRUE}

#you may change the path as necessary

dir<- "D:/cursuri/reproducible research/assignment 1/"    
path<- paste(dir, "","activity.csv" , sep = "", collapse = NULL) 

activity <- read.csv(path)

#check the data and formats for variables
str(activity)
```

Data should have four variables

2)Calculate the total number of steps taken per day

```{r, echo=TRUE, warning=FALSE, message=FALSE}
byday<-summaryBy(steps~date,  data=activity, FUN=sum)
```

do a histogram of the data

```{r, echo=TRUE}
hist(byday$steps.sum, 
     main="Histogram for Total Steps per Day",xlab="Steps", ylab="Days", ylim=c(0,35), 
     xlim=c(0,25000),las=1, breaks=7)
```

compute mean and median of steps by day

```{r, echo=TRUE, warning=FALSE}
stepsperDayMean <- mean(byday$steps.sum, na.rm=TRUE)
stepsperDayMedian <- median(byday$steps.sum, na.rm=TRUE)
```


The average number of steps by day is `r myround(stepsperDayMean, 2) `, and the median number of steps by day is `r stepsperDayMedian `

3) Plot average number of steps for each time interval
```{r, echo=TRUE}
byinterval<-summaryBy(steps~interval,  data=activity, FUN=mean,na.rm=TRUE)
```

```{r, echo=TRUE}
plot(byinterval,  type = "l", main = "Average Number of Steps by Time Interval", xlab = "Time interval",  ylab = "average number of steps")
```




and get the time interval with most steps

```{r, echo=TRUE}
maxinterval <- subset(byinterval, steps.mean==max(byinterval$steps.mean), select=interval)
```
The time interval for which the maximum average number of steps is recorded is `r maxinterval$interval `.

4) impute missing values


```{r, echo=TRUE}
missing <- subset(activity, is.na(activity$steps))
missno<-nrow(missing)
```
The number of missing values is `r missno`.

The strategy chosen is to impute missing values based on the average number of steps per time interval.The new dataset is named 'imputted'

```{r, echo=TRUE}
imputeddata<-merge(activity, byinterval, by.x="interval", by.y="interval", all = TRUE)
imput1<-imputeddata[complete.cases(imputeddata$steps),]
imput2<-imputeddata[!complete.cases(imputeddata$steps),]
imput2$steps <- imput2$steps.mean
imputted<- rbind(imput2,imput1)
```

draw the histogram of steps by day with imputted data

```{r, echo=TRUE}
bydayi<-summaryBy(steps~date,  data=imputted, FUN=sum)
hist(bydayi$steps.sum, 
     main="Histogram for Total Steps per Day with Imputted Values ", xlab="Steps", ylab="Days",
     ylim=c(0,35),xlim=c(0,25000),las=1, breaks=7)
```

compute mean and median of steps by day with imputted data

```{r, echo=TRUE, warning=FALSE}
stepsperDayMeanI <- mean(bydayi$steps.sum, na.rm=TRUE)
stepsperDayMedianI <- median(bydayi$steps.sum, na.rm=TRUE)

```


The average number of steps by day with imputted data is `r myround(stepsperDayMeanI, 2) `, and the median number of steps by day is `r myround(stepsperDayMedianI, 2) `.

This compares with is `r myround(stepsperDayMean, 2) `, and the median number of steps by day is `r stepsperDayMedian ` for original data.

Imputation did not significantly change mean and median, but by comparing thi histograms we see that there is a higher numer of days with total number of steps between 10000 and 15000. 


5) differences between activity patterns in weekdays vs weekends

```{r, echo=TRUE, warning=FALSE, message=FALSE}
#format the date
imputted$dates<- as.Date(imputted$date, format = "%Y-%m-%d")
imputted$days<-weekdays(imputted$dates)
imputted$weekpart<-ifelse(imputted$days==c("Saturday", "Sunday"),"weekend","weekday")

byintervalwd<-summaryBy(steps~interval+weekpart,   data=imputted, FUN=mean,na.rm=TRUE)

xyplot(byintervalwd$steps.mean ~ byintervalwd$interval|byintervalwd$weekpart, main="Average Number of Steps by Time Interval",xlab="Time interval", ylab="Average number of steps",layout=c(2,1), type="l",  aspect = "fill")
```

There is an early peak on weekdays, and more activity on weekends that is more evenly spread and lasts until late hours.

Acknowledgment:

I have used supplementary resources as follows:

- http://stackoverflow.com/

- https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf

- http://www.r-bloggers.com

- http://www.statmethods.net/ (Quick-R by Rob Kabacoff)

- http://kbroman.org/knitr_knutshell/pages/Rmarkdown.html

